Data Management and Domain Adjusting Strategies for Implementing Parallel SPH Method in JPUE Simulation

### 2016-01-06 Paper for submission.tex
fixed 1 2.1

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Zhixuan Cao, Abani Patra}
\IEEEauthorblockA{Department of Mechanical and Aerospace Engineering\\
 The University at Buffalo\\
 Buffalo, New York 14260\\
Email: zhixuanc@buffalo.edu, abani@buffalo.edu}
%\and
%\IEEEauthorblockN{Homer Simpson}
%\IEEEauthorblockA{Twentieth Century Fox\\
%Springfield, USA\\
%Email: homer@thesimpsons.com}
%\and
%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
%\IEEEauthorblockA{Starfleet Academy\\
%San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212\\
%Fax: (888) 555--1212}
}
% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
This paper presents a parallel implementation of smoothed particle hydrodynamics (SPH) method using the message passing interface (MPI) standard to simulate a turbulent jet or plume which is ejected from a nozzle into a uniform environment(JPUE). Background grid is used to reduce neighbors searching cost and decompose (redecompose) the domain. Space Filing Curve(SFC) based index is adopted to assign an unique identifier to each background grid. As simulation of JUPE requires adding of new particles during simulation, time-dependent SFC based indexes are assigned to particles to guarantee uniqueness of identifiers.  
%to distinguish particles added at the same position but different simulation time step. 
Both particle and background grid are managed by hashtables which ensure quick and flexible accessing, adding and deleting of element. Additional link list attached to the hashtable is used to handle hash confliction. A SFC based three dimensional domain decomposition and a dynamic load balancing strategy are implemented to ensure good load balance. In addition, computational domain is adjusted during simulation to reduce computational cost. 
Numerical tests show that our code has acceptable strong scalability and good weak scalability.
These strategies can be further applied to many other implementations of meshfree methods, especially those implemetations that require flexible accessing, adding and deleting of particles and backgroud grids.
\end{abstract}

@@

SPH is a meshless scheme, initially developed for astrophysical applications by Lucy\cite{lucy1977numerical} and Gingold and Monaghan\cite{gingold1977smoothed}. 

@@

A lot of research has been done in parallellization of SPH. 

@@

Chen Wenbo\cite{wenbo2014performance} presented a parallel SPH implementation on multi-core CPUs. 
David W. Holmes\cite{holmes2011framework} presented a simulation framework that enables distributed numerical computing in multi-core shared-memory environments. 
Domínguez\cite{dominguez2011optimization} presented optimizations for both CPU and GPU parallelization of a SPH method. 
Angela Ferrari\cite{ferrari2009new} parallelized a 3D SPH code using the message passing interface (MPI) standard, together with a dynamic load balancing strategy to improve the computational efficiency of the scheme. 
D. Kumar and Abani\cite{kumar2013parallel} implemented CPU parallel Godunov SPH in simulation of granular flows.
A.J.C. Crespo\cite{crespo2015dualsphysics} used the parallel power computing of GPUs to accelerate DualSPHysics by up to two orders of magnitude compared to the performance of the serial version.\\
However, most implementation of parallel SPH method is limitted to bechmark problems like dam break, or relative simple scenarios like breaking-waves, floodings ect. Rare work has been done in simulation of more complicated problems, such as eruption of volcano plume which is essentially a mutilple phases, turbulent, ejection mixing process accompanied by microphysics phenomena like phase change of water, aggregation, reaction ect. Prediction of such more complicated phenomena with acceptable accuracy at given time window can not been acomplished without parallel computing. What's more, imposing of some types of boundary conditions (such as realistic wind feild, eruption boundary condition) requires dynamiclly adding and removing of particles during simulation. This requies efficient and more flexible data management scheme. What need to point out is that the benefit of flexible data management strategies are not only limited to specific implementations of SPH. Actually, flexibility in data management are more desperately needed in some advanced techniques of SPH, such as dynamic particle splitting techniques\cite{vacondio2012accurate, feldman2007dynamic}, which will give greater resolution at the area of interest by spliiting one large particle to several smaller ones.

@@

Fortunately, efficient and flexible data management strategies for high performance computing have been successfully implemented in mesh based methods(eg. \cite{laszloffy2000simple} for adaptive hp FEM, and \cite{pitman2003computing, patra2005parallel} for FVM). Motivated by techniques developed for mesh based methods, we present a complete framwork for parallelizing SPH program with MPI standard model allowing more flexible and efficient data access in this paper.\\
Any implementation of SPH code requires efficient searching and updating of neighbors during simulation. We adopted background grid which was proposed by Monaghan and Lattanzio\cite{monaghan1985refined} and implemented widespreadly in parallel SPH. The background grid is also used for domain decomposition in SPH. We refer to the elements of background grid, namely a squares for two dimension and cubics for three dimension, as buckets. As for the storage of physical quantities associated to each particle, different strategies was adopted in existing implementations of SPH. 
In both SPHysics\cite{dominguez2011optimization} and DualSPHysics\cite{crespo2015dualsphysics}, The physical quantities of each particle (position, velocity, density…) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1)access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome. Angela Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contains all data of particle and bucket. As for the management of data, we adopt an hash table to store pointers to particles and buckets, which gives us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier. The SFC based numberring strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers. As for domain decomposion, even though more complicated graph-based partition tools\cite{biswas1999experiments} might get higher quality decomposition, they requires much more effort in programming. So we adopt an easy-programming scheme based on SFC \cite{patra1999efficient}.\\
To the best of the author's knowledge, no implementation of SPH has the feature of adjusting computational domain based on simulation. For JPUE simulation, such feature will greatly reduce computational cost by avoiding computing of uninfluenced fluid. This feature is accomplished by adding a scan function to monitor the most outside layer of the domain and turn ghost particles to real particles at proper time. 

@@

SPH is a meshfree, Lagrangian method. The domain is discretized by particles and the position of particle is updated at every time step. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of partial differential equations need to be transformed into the Lagrangian formalism. Using a kernel function that provides the weighted estimate of the field variables at a discrete point (particle), the integral equations are evaluated as sums over neighbor particles. Only particles located within support of kernel function will interact. Thus, physical proterties (position, density, velocity, internal energy, pressure) are updated based on its neighbors. So a neighbor searching need to be carried out before updating of physical proterties. We use buckets which overlap the whole domain and keep fixed in time during the entire simulation, to reduce searching cost. Domain decomposition will be based on SFC going through centroids of all buckets. A basic works flow of our SPH code is shown in figure \ref{fig:Work_flow}.\\

@@

We will describe in this section these strategies which satisfies these demand.

@@

SFC-based indexing scheme can guarantee uniqueness of particle identifier only in simple scenarios when particles are added once while setting up initial condition. In some situations, new particles need to be added while simulation. For example, new particles need to be added at the bottom of the eject vent for JPUE simulation. To distinguish particles added at the "same place" (the small area, all points whithin which will be mapped to the same $k$.) at different time steps, we extend the SFC-based key to time-dependent SFC based key by including date of birth of particles into the key. The time-dependent SFC based key can be written as: $[k,t]$, where $t$ is the time step. The map $h_n$ will become:

@@

\item This type of key effectivelly generate a global address space. Globality of key and conservation of locality make it easy to partition the sorted key squence and obtain the decompositiion of the problem.

@@

What need emphasize here is that motion of particles might mess up locality that established based on initial coordinates of particles. As particles are moving regularly, the locality of most of particles should still be conserved during simulation.

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Zhixuan Cao, Abani Patra, Matt Jones, ...}
\IEEEauthorblockA{Department of Mechanical and Aerospace Engineering\\
 The University at Buffalo\\
 Buffalo, New York 14260\\
Email: zhixuanc@buffalo.edu, abani@buffalo.edu}
%\and
%\IEEEauthorblockN{Homer Simpson}
%\IEEEauthorblockA{Twentieth Century Fox\\
%Springfield, USA\\
%Email: homer@thesimpsons.com}
%\and
%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
%\IEEEauthorblockA{Starfleet Academy\\
%San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212\\
%Fax: (888) 555--1212}
}
% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
This paper presents a parallel implementation of smoothed particle hydrodynamics (SPH) method using the message passing interface (MPI) standard to simulate a turbulent jet or plume which is ejected from a nozzle into a uniform environment(JPUE). Background grid is used to reduce neighbors searching cost and decompose (redecompose) the domain. Space Filing Curve(SFC) based index is adopted to assign an unique identifier to each background grid. As simulation of JUPE requires adding of new particles during simulation, time-dependent SFC based indexes are assigned to particles to guarantee uniqueness of identifiers.  
%to distinguish particles added at the same position but different simulation time step. 
Both particle and background grid are managed by hashtables which ensure quick and flexible accessing, adding and deleting of element. %Additional link list attached to the hashtable is used to handle hash conflicts.
 A SFC based three dimensional domain decomposition and a dynamic load balancing strategy are implemented to ensure good load balance. In addition, computational domain is adjusted during simulation to reduce computational cost. 
Numerical tests show that our code has acceptable strong scalability and good weak scalability.
These strategies can be further applied to many other implementations of meshfree methods, especially those implementations that require irregular
particle distributions, adding and deleting of particles and leverage adaptive background grids.
\end{abstract}

@@

SPH (spherical particle hydrodynamics) is a meshless scheme, initially developed for astrophysical applications by Lucy\cite{lucy1977numerical} and Gingold and Monaghan\cite{gingold1977smoothed}. 

@@

A lot of research has appeared  on  parallellization of SPH in the last few years. 

@@

 Wenbo\cite{wenbo2014performance} presented a parallel SPH implementation on multi-core CPUs. 
  Holmes\cite{holmes2011framework} presented a simulation framework that enables distributed numerical computing in multi-core shared-memory environments. 
Dominguez\cite{dominguez2011optimization} presented optimizations for both CPU and GPU parallelization of a SPH method. 
 Ferrari\cite{ferrari2009new} parallelized a 3D SPH code using the message passing interface (MPI) standard, together with a dynamic load balancing strategy to improve the computational efficiency of the scheme. 
Kumar et. al. \cite{kumar2013parallel} implemented a parallel Godunov SPH in simulation of granular flows.
 Crespo\cite{crespo2015dualsphysics} used the GPUs to accelerate DualSPHysics by up to two orders of magnitude compared to the performance of the serial version.\\
However, most implementation of parallel SPH method presented to date are limited to standard SPH and benchmark problems like dam break, or relatively
 simple scenarios like breaking-waves, floodings etc. Work on more complicated problems, such as eruption of volcano plume which is essentially a mult-phase turbulent flow,
  ejection mixing process accompanied by microphysics phenomena like phase change of water, aggregation, reaction etc. is relatively rare. 
  Prediction of such  c complicated phenomena with acceptable accuracy at given time window requires resolutions (very high particle counts) that cannot been accomplished without parallel computing. What's more, imposing of some types of boundary conditions (such as realistic wind field, eruption boundary condition) requires dynamically adding and removing of particles during simulation. This requires efficient and more flexible data management scheme. What we need to point out is that the benefit of flexible data management strategies are not only limited to specific implementations of SPH. Actually, flexibility in data management are more critical for several advanced techniques of SPH, such as dynamic particle splitting techniques\cite{vacondio2012accurate, feldman2007dynamic}, which will give greater resolution at the area of interest by splitting one large particle to several smaller ones.

@@

Fortunately, efficient and flexible data management strategies for high performance computing have been successfully implemented in mesh based methods (eg. \cite{laszloffy2000simple} for adaptive hp FEM, and \cite{pitman2003computing, patra2005parallel} for FVM). Motivated by techniques developed for mesh based methods, we present a complete framwork for parallelizing SPH program with MPI standard model allowing more flexible and efficient data access in this paper.\\
Any implementation of SPH code requires efficient searching and updating of neighbors during simulation. Of the many choices possible for this we adopted a background grid which was proposed by Monaghan and Lattanzio\cite{monaghan1985refined} and is quite popular in parallel SPH. The background grid is also used for domain decomposition in SPH. We refer to the elements of background grid, namely  squares for two dimension and cubics for three dimension, as buckets.

As for the actual storage of data representing the physical quantities associated to each particle, different strategies  have been adopted in existing implementations of SPH. 
In both SPHysics\cite{dominguez2011optimization} and DualSPHysics\cite{crespo2015dualsphysics}, The physical quantities of each particle (position, velocity, density…) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome.   Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contains all data of particle and bucket. As for the management of data, we adopt a hash table to store pointers to particles and buckets, which gives us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers.

As for domain decomposition, even though more complicated graph-based partition tools\cite{biswas1999experiments} might get higher quality decomposition, they requires much more effort in programming and computation. So we adopt an easy-programming scheme based on SFC \cite{patra1999efficient}.\\
To the best of the author's knowledge, no implementation of SPH has the feature of adjusting computational domain based on simulation needs. For JPUE simulation, such feature will greatly reduce computational cost by avoiding computing of uninfluenced fluid. This feature is accomplished by adding a scan function to monitor the outermost layer of the domain and turn ghost particles to real particles at the proper time. 

@@

SPH is a meshfree, Lagrangian method. The domain is discretized by particles and the position of each particle is updated at every time step. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of partial differential equations need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimate of the field variables in the neighborhood of a discrete point (particle), the integral equations are evaluated as sums over neighbor particles. Only particles located within support of kernel function will interact. Thus, physical properties (position, density, velocity, internal energy, pressure) are updated based on its neighbors. So a neighbor search needs to be carried out before updating of physical properties. We use buckets which contain all particles associated with a sub-domain 
 and are kept fixed in time during the entire simulation, to reduce search cost (since search can now be restricted to only neighboring domains).
  Domain decomposition will be based on an SFC going through centroids of all buckets. A basic work flow of our SPH code is shown in figure \ref{fig:Work_flow}.\\

@@

We will describe in this section   strategies which satisfies these requirements.

@@

SFC-based indexing scheme can guarantee uniqueness of particle identifier only in simple scenarios when particles are added once while setting up initial condition. In some situations, new particles need to be added while simulation. For example, new particles need to be added at the bottom of the eject vent for JPUE simulation. To distinguish particles added at the ``same place" (the small area, all points within which will be mapped to the same $k$.) at different time steps, we extend the SFC-based key to time-dependent SFC based key by including date of birth of particles into the key. The time-dependent SFC based key can be written as: $[k,t]$, where $t$ is the time step. The map $h_n$ will become:

@@

\item This type of key effectively generates a global address space. Globality of key and conservation of locality make it easy to partition the sorted key sequence and obtain a decompositiion of the problem.

@@

What we need to emphasize here is that in theory motion of particles will destroy locality established based on initial coordinates of particles. However, as particles are moving fairly regularly in a JPUE, the locality of most of particles should still be conserved during simulation.
